Expos\'{e} Master Thesis\\ Checking Assertions for Smart Contracts

### 2020-10-20 meeting-notes/20200928-expose.tex
examples for syntax

@@

As an example, consider a contract that takes a prime number as a parameter.
\begin{lstlisting}[numbers=none]
contract Example {
  function (int p) public {
    // assume p is prime
    ...
  }
}
\end{lstlisting}
This assumption can be expressed with an explicit assertion in predicate logic.
\begin{displaymath}
  (\forall n) (2 \le n \le \sqrt p) \Rightarrow (p \mathbin{\%} n) \ne 0
\end{displaymath}
To test the validity of this assumption requires a loop in the contract, but the test would take
$O(\sqrt p)$ time (assuming constant time for computing the remainder) and thus produce extra cost
linear in $\sqrt p$ accordingly. 
However, we could do better by recruiting the validators of the contract for a distributed
effort to find a counterexample. To this end, we consider the negation of the assertion.
\begin{gather}\label{eq:4}
  (\exists n) (2 \le n \le \sqrt p) \wedge (p \mathbin{\%} n) = 0
\end{gather}
This assertion can be checked pointwise by having each validator independently choose a
random $n$ fulfilling $2 \le n \le 
\sqrt p$ and checking whether $(p \mathbin{\%} n) = 0$. If the remainder is $0$, the
validator found a counterexample, posts its veto to the P2P net, and stops further
exection. Otherwise, it accepts $p$ knowing that other points will be checked by other
validators.

@@

\begin{itemize}
\item Formally define a set of logical formulas that are amenable to the kind of checking
  described above. In a first approximation, the interesting assertion formulas seem to
  be those with prenex universal quantification. 
\item Collect examples for interesting formulas.
\item Formalize the cost incurred by checking a formula.
\item Extend the contract language with assertions stating these logical
  formulas. There are several approaches to perform this extension.
  \begin{enumerate}
  \item (Solidity) Extend the syntax of the assert statement. Requires an extension of
    the compiler.
  \item (Tezos/Solidity) The assertion is stated in its own syntax in a file separate from the
    contract implementation. Requires a dedicated compiler of assertions to
    Michelson. The generated code is either combined with the contract implementation
    or stored separately on the blockchain. Several entry points may lead to difficulties.
  \end{enumerate}
\item Implement the transformation in two steps:
  \begin{enumerate}
  \item negation of the formula
  \item followed by the translation of an existentially quantified formula to code that picks a
    random value subject to predicate. For initial experiments, the target of the
    translation can be any programming 
    language; later on, the target has to be a language hosted on the blockchain
    (e.g. Michelson or Solidity). These languages have to be extended with a random
    generator. There has to be a special execution mode for ``challenge contracts''
    that search for a counterexample of an assertion.
  \end{enumerate}
  Negation turns prenex universial quantification into prenex existential quantification. But it
  would be ineffective to generate a random integer and then check that it falls in the
  interesting range (e.g., $2\le n \le \sqrt p$ and $0\le k< n-1$). Rather, the formula
  needs to be analzed and the random generator should be restricted to only generate
  interesting values.

@@

>>>>

\usepackage{hyperref}

@@

As an example, consider a contract that takes a prime number as a parameter.
\begin{lstlisting}[numbers=none]
contract Example {
  function (int p) public {
    // assume p is prime
    ...
  }
}
\end{lstlisting}
This assumption can be expressed with an explicit assertion in predicate logic.
\begin{gather}\label{eq:5}
  (\forall n) (2 \le n \le \sqrt p) \Rightarrow (p \mathbin{\%} n) \ne 0
\end{gather}
To test the validity of this assumption requires a loop in the contract, but the test would take
$O(\sqrt p)$ time (assuming constant time for computing the remainder) and thus produce extra cost
linear in $\sqrt p$ accordingly. 
However, we could do better by recruiting the validators of the contract for a distributed
effort to find a counterexample. To this end, we consider the negation of the assertion.
\begin{gather}\label{eq:4}
  (\exists n) (2 \le n \le \sqrt p) \wedge (p \mathbin{\%} n) = 0
\end{gather}
This assertion can be checked pointwise by having each validator independently choose a
random $n$ fulfilling $2 \le n \le 
\sqrt p$ and checking whether $(p \mathbin{\%} n) = 0$. If the remainder is $0$, the
validator found a counterexample, posts its veto to the P2P net, and stops further
exection. Otherwise, it accepts $p$ knowing that other points will be checked by other
validators.

@@

\begin{itemize}
\item Formally define a set of logical formulas that are amenable to the kind of checking
  described above. In a first approximation, the interesting assertion formulas seem to
  be those with prenex universal quantification.

  In a first iteration, we want to restrict ourselves to formulas of
  predicate logic with prenex universal quantification only.
  \begin{align*}
    t &::= \text{primitive types like int, bool, string, \dots} \\
    \Theta &::= \Phi \mid \forall (x:t) \Theta \\
    \Phi,\Psi &::= \neg\Phi \mid \Phi\wedge\Psi \mid \Phi\vee\Psi \mid
                M \rho N \\
    \rho &::= < \mid > \mid \le \mid \ge \mid = \mid \ne \mid \dots \\
    M, N &::= x \mid c \mid M \oplus N  \mid f (\overline M) \\
    \oplus &::= +\mid -\mid * \mid / \mid \% \mid \dots \\
    c &::= \text{constants: numbers, strings, etc} \\
    f &::= \text{operations (existing in the blockchain VM)}
  \end{align*}
  Later, we might consider existential quantification and user-defined
  functions (for example, sqrt, length of a list, access list element
  are not predefined by the VM and have to be implemented. The list
  operations would be really good to have, though). See
  Section~\ref{sec:suggestions-tezos} for suggestions for a concrete syntax.
\item Collect examples for interesting formulas.
\item Formalize the cost incurred by checking a formula.
\item Extend the contract language with assertions stating these logical
  formulas. There are several approaches to perform this extension.
  \begin{enumerate}
  \item (Solidity) Extend the syntax of the assert statement. Requires an extension of
    the compiler.
  \item (Tezos/Solidity) The assertion is stated in its own syntax in a file separate from the
    contract implementation. Requires a dedicated compiler of assertions to
    Michelson. The generated code is either combined with the contract implementation
    or stored separately on the blockchain. Several entry points may lead to difficulties.
  \end{enumerate}
\item Implement the transformation in two steps:
  \begin{enumerate}
  \item negation of the formula
  \item followed by the translation of an existentially quantified formula to code that picks a
    random value subject to predicate. For initial experiments, the target of the
    translation can be any programming 
    language; later on, the target has to be a language hosted on the blockchain
    (e.g. Michelson or Solidity). These languages have to be extended with a random
    generator. There has to be a special execution mode for ``challenge contracts''
    that search for a counterexample of an assertion.
  \end{enumerate}
  Negation turns prenex universial quantification into prenex existential quantification. But it
  would be ineffective to generate a random integer and then check that it falls in the
  interesting range (e.g., $2\le n \le \sqrt p$ and $0\le k< n-1$). Rather, the formula
  needs to be analzed and the random generator should be restricted to only generate
  interesting values.

@@

\section{Suggestions (Tezos)}
\label{sec:suggestions-tezos}

The syntax for assertions should be chosen to be similar to OCaml.
A proposal for the assertion in \eqref{eq:5} could be
\begin{lstlisting}[language=caml]
parameter (_, (_, p)) =
  forall (n : int)
    if (2 <= n && n <= p / 2)
      p % n <> 0
\end{lstlisting}
Such a parameter phrase may occur multiple times. Its single argument
is a pattern that matches the contract's argument. The type of the
pattern variables (here just \lstinline/p/) is determined from the code of the
contract.

Multiple parameter phrases can be used to state assertion for
contracts with multiple entry points. In Tezos/Michelson, entry points are
modeled by having a sum type as parameter type. For example, a
contract with two entry points can be equipped with assertions as
follows.
\begin{lstlisting}[language=caml]
parameter (Left x) =
  0 < x && x < 100

parameter (Right (a, b)) =
  exists (c : int)
    if (0 < c && c <= a + b)
      a*a + b*b = c*c
\end{lstlisting}
BTW, this is an example for a useful contract with an existential!

The idea would be to have two files with the same basename, but
different extensions. One file contains the Michelson code, the other
the assertion.
